---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rpicsa: R Package for PICSA and Climatic Analysis for Crops

<!-- badges: start -->
[![R-CMD-check](https://github.com/IDEMSInternational/rpicsa/workflows/R-CMD-check/badge.svg)](https://github.com/IDEMSInternational/rpicsa/actions)
[![Codecov test coverage](https://codecov.io/gh/IDEMSInternational/rpicsa/branch/main/graph/badge.svg)](https://app.codecov.io/gh/IDEMSInternational/rpicsa?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![license](https://img.shields.io/badge/license-LGPL%20(%3E=%203)-lightgrey.svg)](https://www.gnu.org/licenses/lgpl-3.0.en.html)
<!-- badges: end -->

## Overview

**rpicsa** is an R package designed to assist in the analysis of climate data for crop production using the Plant Impact on Climate and Societal Adaptation (PICSA) framework. It provides a range of functions to help researchers, agronomists, and farmers make informed decisions about planting and managing crops based on climatic data.

The package offers a suite of tools to perform various climatic analyses, including:

- Start of Rains: Determine the onset of the rainy season, an important factor in crop planning.
- End of Rains: Identify when the rains conclude, helping in crop harvesting and preparation for the dry season.
- End of Season: Identify when the rainy season concludes, helping in crop harvesting and preparation for the dry season.
- Length of Season: Calculate the duration of the growing season, a useful parameter for crop selection.
- Rainfall Summaries: Generate summaries of rainfall within a rainy season to understand precipitation variability.
- Temperature Summaries: Calculate temperature statistics to assess the impact on crop development.
- Crop Success Probability: Estimate the likelihood of successful crop production based on historical climate data.
- Season Start Probability: Predict the probability of the rainy season's onset to optimise planting schedules.

## Installation

You can install the development version of `rpicsa` like so:

```{r, eval = FALSE}
# Install the 'devtools' package if you haven't already
if (!require(devtools)) {
  install.packages("devtools")
}

# Install 'rpicsa' from GitHub
devtools::install_github("IDEMSInternational/rpicsa")
```

## Usage

We give some small examples in using the functions in `rpicsa`

```{r, message = FALSE, warning = FALSE}
library(rpicsa)
library(dplyr)

# use the data in the package, but for this demonstration we will look at the data before 1950.
daily_niger <- rpicsa::daily_niger %>% dplyr::filter(year <= 1950)
```

To calculate the start of rains in Niger, we can use the `start_rains()` function. In our case, we define the start of rains as the first instance where there is more than 25mm of rainfall over three days, and there is no more than nine days without rainfall within 21 days.

We also calculate the end of rains. Here, we define the end of rains to be the first instance after day 121 (April 30th), where there is one day with no more than 10mm of rainfall.

```{r, message = FALSE, warning = FALSE, eval=FALSE, include=TRUE}
# Calculate Start of Rains
start_rain_table <- start_rains(data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain",
            total_rainfall = TRUE, over_days = 3, amount_rain = 25,
            dry_spell = TRUE, spell_interval = 21, spell_max_dry_days = 9
            )

# Calculate End of Rains
end_rain_table <- end_rains(data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain",
            start_day = 121, interval_length = 1, min_rainfall = 10)

# Combine into a table Start of Rains
summary_rains_table <- dplyr::full_join(start_rain_table, end_rain_table)
```

We can then use this to calculate other summary statistics - for example, length of season, or rain summaries in the season

```{r, message = FALSE, warning = FALSE, eval=FALSE, include=TRUE}
# Length of season
season_length_table <- rpicsa::seasonal_length(summary_rains_table, start_date = "start_rains", end_date = "end_rains",
                                               data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain")

# We can also look at the total rainy days and rainfall amount in each season
season_rain_table <- rpicsa::seasonal_rain(summary_rains_table, start_date = "start_rains", end_date = "end_rains",
                                               data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain")

```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Calculate Start of Rains
start_rain_table <- start_rains(data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain",
            total_rainfall = TRUE, over_days = 3, amount_rain = 25,
            dry_spell = TRUE, spell_interval = 21, spell_max_dry_days = 9
            )

# Calculate End of Rains
end_rain_table <- end_rains(data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain",
            start_day = 121, interval_length = 1, min_rainfall = 10)

# Combine into a table Start of Rains
summary_rains_table <- dplyr::full_join(start_rain_table, end_rain_table)
# Length of season
season_length_table <- rpicsa::seasonal_length(summary_rains_table, start_date = "start_rains", end_date = "end_rains",
                                               data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain")

# We can also look at the total rainy days and rainfall amount in each season
season_rain_table <- rpicsa::seasonal_rain(summary_rains_table, start_date = "start_rains", end_date = "end_rains",
                                               data = daily_niger, date_time = "date", station = "station_name",
            year = "year", doy = "doy", rain = "rain")
# Bind together to view
summary_rains_table <- dplyr::full_join(summary_rains_table, season_length_table)
dplyr::full_join(summary_rains_table, season_rain_table) %>%
  knitr::kable()
```